---
title: "Homework 4"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due:** Friday, September 25, 11:59pm on Canvas

**Instructions:** 

1. Go to Canvas -> Assignments -> HW 4. Open the GitHub Classroom assignment link
2. Follow the [instructions](https://sta261-f26.github.io/resources/github_instructions/) to accept the assignment and clone the repository to your local computer
3. The repository contains the file `hw_04.qmd`. Write your code and answers to the questions in the Quarto document. Commit and push to GitHub regularly.
4. When you are finished, make sure to Render your Quarto document; this will produce a `hw_04.md` file which is easy to view on GitHub. Commit and push both the `hw_04.qmd` and `hw_04.md` files to GitHub
5. When your code generates figures (e.g., in question 6), you will also need to commit and push those figures to GitHub. These figures will be stored in a folder called something like `homework_4_files`, in the same directory as your `.qmd` and `.md` files. When you Git commit your work, **add** the image files *as well* as the `.qmd` and `.md` files.
6. Finally, [request feedback](https://sta261-f26.github.io/resources/github_instructions/#step-10-request-feedback) on your assignment on the "Feedback" pull request on your HW 4 repository

**Important:** Make sure to include the `.qmd` and `.md` files, and any images, when you submit!

**Collaborators and AI statement:** At the top of your assignment, list any collaborators (Dr. Evans does not count -- he is a resource, not a collaborator), and provide an AI statement describing your use of AI to complete the questions. If you did not use AI in any way, let me know that too. If you did use AI, provide: the model used, the HW questions on which you used AI, and a summary of the ways in which AI provided assistance. As described in the syllabus, AI use is permitted on homework assignments, but submissions without a complete AI statement may not receive full credit.

**Code guidelines:**

* If a question requires code, and code is not provided, you will not receive full credit
* You will be graded on the quality of your code. In addition to being correct, your code should also be easy to read
* Some questions require both code and discussion of the results. If you are missing a discussion of the results, you will not receive full credit

# Exam practice

Because this is a computing course, exams will involve reading and writing small pieces of code and interpreting output. As it is not feasible for these to be done electronically, you will be expected to write small portions of code by hand. 

The amount you will need to do by hand is not as extensive as the amount of code on a homework assignment! However, you will need to be familiar with the fundamental types of functions and operations we have used in the course.

To help you practice those skills, I have included some sample questions on this homework assignment that are similar to some of the questions I could ask on an exam. I suggest that you try **writing these by hand** on a piece of paper, before typing the answers.

## Writing code

In each of the questions below, write code to produce the output from the original data. Note that you cannot actually run this code, since you don't have easy access the original data! That is ok. Use `eval = FALSE` in your chunk options to show the code but prevent it running.

:::{.question}
#### Question 1

Original data:

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(palmerpenguins)

set.seed(3)

sub_data <- penguins |>
  slice_sample(n=10) |>
  select(species, island, bill_length_mm, bill_depth_mm)

sub_data
```

Output:

```{r, echo=F}
sub_data |>
  count(species, island)
```

:::

:::{.question}
#### Question 2

Original data:

```{r, echo=F}
ex_df <- data.frame(
  id = c(1, 2, 3),
  x_1 = c(3, 1, 4),
  x_2 = c(5, 8, 9),
  y_1 = c(0, 1, 2),
  y_2 = c(2, 7, 9)
)

ex_df
```

Output:

```{r, echo=F}
ex_df |>
  pivot_longer(cols = -id, names_to = c("group", "obs"), names_sep = "_")
```
:::

:::{.question}
#### Question 3

Original data:

```{r, echo=F}
ex_df <- data.frame(
  id = c(1, 1, 2, 2, 3, 3),
  group = c("x", "y", "x", "y", "x", "y"),
  value = sample(1:6, replace=T)
)

ex_df
```

Output:

```{r, echo=F}
ex_df |>
  pivot_wider(id_cols = id, names_from = group, values_from = value)
```
:::

:::{.question}
#### Question 4

Here there are two original datasets (`df1` and `df2`):


```{r, echo=F}
df1 = data.frame(
  a_x = 1,
  a_y = 2,
  b_x = 2,
  b_y = 3
)

df2 = data.frame(
  id = c("a", "b"),
  z = c(4, 5)
)
```

```{r}
df1

df2
```

Output:


```{r, echo=F}
df1 |> 
  pivot_longer(cols = -c(), names_to = c("id", ".value"), names_sep = "_") |>
  left_join(df2, join_by(id))
```
:::


:::{.question}
#### Question 5

Create a list `x` in R such that:

* `x[[1]]` returns the function `mean`
* `x[[2]]` returns the function `sd`
* `x[[3]][[1]]` returns the vector `c(0, 1, 2)`
* `x[[3]][[2]]` returns a function which calculates the cube root of a vector

:::

# Simpsons data

## Motivation

*The Simpsons* is a long-run animated sitcom which has had a big effect on popular culture over the years. The data for this assignment is composed of four files that contain the characters, locations, episode details, and script lines for approximately 600 Simpsons episodes between 2010 and 2016. 

### Data

The code below will download the different datasets:

```{r, eval = F}
simpsons_characters <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-04/simpsons_characters.csv')
simpsons_episodes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-04/simpsons_episodes.csv')
simpsons_locations <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-04/simpsons_locations.csv')
simpsons_script_lines <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-04/simpsons_script_lines.csv')
```

### `simpsons_characters.csv`

| variable        | class     | description                                 |
|:----------------|:----------|:-------------------------------------------|
| id              | double    | Unique identifier for each character record.          |
| name            | character | Full name associated with the character record.       |
| normalized_name | character | Lowercase version of the character name.           |
| gender          | character | Gender associated with the character record.          |

### `simpsons_episodes.csv`

| variable               | class     | description                                         |
|:-----------------------|:----------|:---------------------------------------------------|
| id                     | double    | Unique identifier for each episode record.                 |
| image_url              | character | URL linking to the image associated with the episode record. |
| imdb_rating            | double    | IMDb rating for the episode. |
| imdb_votes             | double    | Number of votes received on IMDb for the episode. |
| number_in_season       | double    | Episode number within the season. |
| number_in_series       | double    | Episode number within the series. |
| original_air_date      | date      | Date the episode originally aired. |
| original_air_year      | double    | Year the episode originally aired. |
| production_code        | character | Code used in production to identify the episode. |
| season                 | double    | Season number of the episode. |
| title                  | character | Title of the episode. |
| us_viewers_in_millions | double    | Number of viewers in the U.S. in millions. |
| video_url              | character | URL linking to the video associated with the record. |
| views                  | double    | Total number of views recorded for the episode video URL. |

### `simpsons_locations.csv`

| variable        | class     | description                                 |
|:----------------|:----------|:-------------------------------------------|
| id              | double    | Unique identifier for each location.        |
| name            | character | Name of the location.                       |
| normalized_name | character | Lowercase version of the location name.  |

### `simpsons_script_lines.csv`

| variable           | class     | description                                              |
|:-------------------|:----------|:--------------------------------------------------------|
| id                 | double    | Unique identifier for each script line. |
| episode_id         | double    | Identifier for the episode in which the line appears. |
| number             | double    | Sequential number of the line within the episode. |
| raw_text           | character | The original text of the script line. |
| timestamp_in_ms    | double    | Timestamp of the line in milliseconds. |
| speaking_line      | logical   | Indicates whether the line is spoken by a character. |
| character_id       | double    | Identifier for the character speaking the line. |
| location_id        | double    | Identifier for the location where the line is spoken. |
| raw_character_text | character | Original text of the character's name. |
| raw_location_text  | character | Original text of the location name. |
| spoken_words       | character | Words spoken by the character in the line. |
| normalized_text    | character | Lowercase version of the script line. |
| word_count         | double    | Number of words in the line. |

## Questions of interest

* Which character has the most spoken lines across all episodes, and how has their dialogue volume changed over the seasons?
* What are the most frequently used locations in the series, and do specific locations correspond to higher IMDb ratings for episodes?
* Is there a relationship between the number of U.S. viewers (in millions) and the IMDb ratings or votes for episodes?

## Your task

1. Examine the data in R, and make sure you understand what information is contained in each dataset.
2. **Before you do any data wrangling or write any code**, brainstorm some ideas for how you will answer the questions of interest. What data wrangling steps might be involved? How would you quantify the questions of interest? Do you need to combine or reshape any of the datasets? (You do **not** need to submit this brainstorming, but it is good practice to think about your analysis before you start writing the code.)
3. Use the data wrangling tools in `R` we have learned in this class to implement your ideas from (2) and explore the questions of interest.

## What you will submit

:::{.question}
#### Question 6

Submit a summary of your analysis that answers the three questions of interest. The summary must meet the following requirements:

* All questions must be addressed
* At least one visualization must be made for each question
* If a statistical model would be appropriate to answer a research question, it should be used
* All results (summary statistics, visualizations, models, etc.) must be discussed
* Discussion does not have to be long, but it should answer the questions and be written in full sentences
:::


**Acknowledgment:** This assignment is borrowed from the [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-03-10/readme.md) challenge from February 4, 2025, which was curated by Nicolas Foss, Prashant Banerjee, and Todd W. Schneider.

