---
title: "Intro to Python"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

## Course so far

**So far**, we know how to do key wrangling tasks in R:

* Grouping, summarizing, mutating
* Pivoting and joining

**What is next:** Data wrangling in Python!

## What *is* R?

* R is a programming language specifically designed for statistics and data analysis
  * Objects for storing data, and functions for interacting with data, are fundamental
  * R is very good at graphics and visualization
  * R is easily extended. Users can write and share their own functions and packages
* We can interact with R through IDEs like RStudio

## What other options exist?

* SAS
* Stata
* SPSS
* Excel
* Python
* Julia
* Matlab
* Many others...

## What is Python

* Python is a general-purpose programming language 
* Like R, python has a wide range of packages to extend functionality
* Certain Python packages allow for sophisticated data analysis and modeling
  * `SciPy`, `NumPy`
  * `scikit-learn`, `statsmodels`, `pytorch`
  * `pandas`
  * `matplotlib`
  
## R vs. Python

My own, *personal*, preferences:

:::: {.columns}

::: {.column width="50%"}
R is good for

* Data visualization and wrangling
* Classical statistics
* Statistical inference
* New statistical methods
:::

::: {.column width="50%"}
Python is good for

* General-purpose programming
* Challenging data types (e.g. images)
* Prediction and machine learning
:::

::::

## Warmup

What do you think this code is doing? Discuss with your neighbor for 1--2 minutes, then we will discuss as a class.

```{python}
#| eval: false
import pandas as pd
import seaborn as sns

penguins = sns.load_dataset("penguins")

(penguins.groupby('species').
  agg(
    mean_bill = ('bill_length_mm', 'mean'),
    sd_bill = ('bill_length_mm', 'std')
    ))
```

## Warmup

```{python}
import pandas as pd
import seaborn as sns

penguins = sns.load_dataset("penguins")

(penguins.groupby('species').
  agg(
    mean_bill = ('bill_length_mm', 'mean'),
    sd_bill = ('bill_length_mm', 'std')
    ))
```

## Data wrangling ideas

* choose certain rows and columns
* calculate summary statistics
* group rows together
* create new columns
* apply functions across columns
* reshape data by pivoting
* join tables

These ideas are language-agnostic! The implementation is just a bit different

## Basic information

```{python}
penguins
```


## Basic information

```{python}
penguins.shape
```

```{python}
penguins.columns
```

## Choosing columns

```{python}
penguins['bill_length_mm']
```

## Choosing columns

```{python}
penguins[['bill_length_mm', 'bill_depth_mm']]
```

## Calculating summary statistics

```{python}
penguins.agg({'bill_length_mm': 'mean'})
```


```{python}
penguins.agg(mean_length = ('bill_length_mm', 'mean'))
```

## Summarizing multiple columns

```{python}
penguins[['bill_length_mm', 'bill_depth_mm']].agg('mean')
```

```{python}
penguins[['bill_length_mm', 'bill_depth_mm']].agg(['mean', 'std'])
```

## Grouping and summarizing

```{python}
(penguins.groupby('species').
  agg(
    mean_bill = ('bill_length_mm', 'mean'),
    sd_bill = ('bill_length_mm', 'std')
    ))
```

## Grouping and summarizing

```{r, include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r}
penguins |>
  group_by(island) |>
  summarize(mean_length = mean(bill_length_mm))
```

## Grouping and summarizing

```{r}
penguins |>
  group_by(species) |>
  summarize(mean_length = mean(bill_length_mm,
                               na.rm = T))
```

## Sorting

```{python}
(penguins.groupby('species').
  agg(
    mean_bill = ('bill_length_mm', 'mean'),
    sd_bill = ('bill_length_mm', 'std')
    ).
  sort_values('mean_bill'))
```

## Sorting

```{python}
(penguins.groupby('species').
  agg(
    mean_bill = ('bill_length_mm', 'mean'),
    sd_bill = ('bill_length_mm', 'std')
    ).
  sort_values('mean_bill', ascending = False))
```


## Class activity

* Work with a neighbor on the class activity
* You do not need to submit this activity!
