---
title: "Data types in R"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

## Course so far

**So far**, we know how to do key wrangling tasks in R:

* Grouping, summarizing, mutating
* Pivoting and joining

**What is next:** Data wrangling in Python!

**To get there:**

* How does R store data?
* Similarities and differences between R and Python objects for data storage and wrangling

## Atomic vectors

So far, we have worked with rectangular data (data tables which are collections of columns):

```{r, echo=F, message=F, warning=F}
library(tidyverse)
as.data.frame(starwars) |>
  select(name, height, species, mass, hair_color) |> head()
```

## Atomic vectors

So far, we have worked with rectangular data (data tables which are collections of columns):

```{r, echo=F, message=F, warning=F}
library(tidyverse)
as.data.frame(starwars) |>
  select(name, height, species, mass, hair_color) |> head()
```

Each column is an **atomic vector**:

```{r}
starwars$name
```


## Atomic vectors

**Atomic vectors** are one-dimensional objects, and are indexed with `[ ]`:

```{r}
x <- c(0, 23, 98)
length(x)

x[1]
x[2]

x[1:2]
x[c(1, 3)]
```

## Atomic vectors

Vectors can contain numbers, booleans, characters, etc:

```{r, echo=T}
x <- c(0, 1, 2)
x
typeof(x)

x <- c("a", "b", "c")
x
typeof(x)
```

The `typeof` function tells what *type* of object we have

## Vectors of multiple types?

```{r, echo=T, eval=F}
x <- c(0, 1, "a")
x
x[1] + 1
```

What do you think will happen when we run this code?


## Vectors of multiple types?

```{r, echo=T, error=T}
x <- c(0, 1, "a")
x
x[1] + 1
```

## Vectors of multiple types?

```{r, echo=T, error=T}
x <- c(0, 1, "a")
x
x[1] + 1
```

**Atomic vectors** only contain one type.

## Lists

```{r, echo=T}
x <- list(c(0, 1), "a")
x
```

## Lists

```{r, echo=T}
x <- list(c(0, 1), "a")
x

x[[1]]

x[[1]][1]
```


## Lists

```{r, echo=T}
x <- list(c(0, 1), "a")
x

x[[1]]

x[[1]][1]

typeof(x[[1]])

x[[2]]

typeof(x[[2]])
```

## Visualizing list structure

```{r, echo=T}
x1 <- list(c(1, 2), c(3, 4))
x1
x2 <- list(list(1, 2), list(3, 4))
x2
```

## Indexing lists

```{r, echo=T}
x <- list(c(1, 2), c(3, 4))

x[1]
typeof(x[1])

x[[1]]
typeof(x[[1]])
```

* `x[1]` returns a *list* which contains the first component of `x`
* `x[[1]]` returns the object stored in the first component

## Rectangular data 

```{r, message=F, warning=F}
starwars
```

## Rectangular data in R

In R, there are two main ways of storing rectangular data:

* matrices
* data frames

## Matrices

A *matrix* generalizes a vector to *two* dimensions:

```{r}
x <- matrix(c(1, 2, 3, 4, 5, 6), nrow=2)
x
```

* Each row is an atomic vector
* Each column is an atomic vector

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r, eval=F}
x[2, 3]
```

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r}
x[2, 3]
```

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r, eval=F}
x[2,]
```

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r}
x[2,]
```

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r, eval=F}
x[,2]
```

## Indexing matrices

```{r}
x <- matrix(c("a", "b", "c", "d", "e", "f"), nrow=2)
x
```

What will this code return?

```{r}
x[,2]
```

## Uses and limitations of matrices

* Correspond to the matrices we know and love from linear algebra
* Usually the right way to store 2-d data for doing math (like matrix multiplication)
* Like atomic vectors, contain only one type of data

```{r}
x <- matrix(c(1, 2, 3, 'a', 5, 6), nrow=2)
x
```

## Data frames

```{r}
example_df <- data.frame(x = c(1, 2, 3),
                         y = c('a', 'b', 'c'))
example_df
```

## Aside: what *are* data frames?

```{r}
example_df <- data.frame(x = c(1, 2, 3),
                         y = c('a', 'b', 'c'))
typeof(example_df)
```

* Matrices are like a 2-d atomic vector
* Data frames are a special type of list! With some requirements:
  * Each component is a vector
  * Each component has the same length
  
## Indexing data frames

`[ ]` can work for indexing data frames, just like matrices:

```{r}
example_df <- data.frame(x = c(1, 2, 3),
                         y = c('a', 'b', 'c'))
example_df
example_df[2, 1]
```

## Indexing data frames

Like lists, `[[ ]]` and `$` can also be used:

```{r}
example_df <- data.frame(x = c(1, 2, 3),
                         y = c('a', 'b', 'c'))
example_df

example_df$x
example_df[["x"]]
```

## Class activity

* Work with a neighbor on the class activity
* At the end of class, submit your work as an HTML file on Canvas (one per group, list all your names)
