---
title: "Data wrangling fundamentals"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

## Last time

* `filter`: choose certain rows
* `summarize`: calculate summary statistics
* `group_by`: group rows together
* `mutate`: create new columns

## Input data

**Key principle:** The first argument to a `dplyr` function is the input dataframe

```{r, include=F}
library(tidyverse)
```

Example:

```{r, eval=F}
filter(starwars, species == "Droid")
```

```{r, echo=F}
starwars |>
  filter(species == "Droid") |>
  head(2)
```

## Input data and the pipe

By default, the pipe (`|>`) enters the output of one line as the first argument in the next line

```{r, eval=F}
starwars |>
  filter(species == "Droid")
```

<br/>

<br/>

```{r, eval=F}
starwars |>
  filter(species == "Droid") |>
  summarize(mean_height = mean(height))
```

## Data in, data out

**Key principle:** The **input** to a `dplyr` function is always a data frame, and the **output** is always a dataframe

```{r}
starwars |>
  filter(species %in% c("Droid", "Human", "Ewok")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm=T))
```


## The output

**Question:** After running this code, have we changed the original `starwars` data?

```{r}
starwars |>
  filter(species %in% c("Droid", "Human", "Ewok")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm=T))
```

## The output

**Key principle:** `dplyr` functions do **not** modify the original data; they return a *new* data frame

```{r}
starwars |>
  filter(species %in% c("Droid", "Human", "Ewok")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm=T))

dim(starwars)
```

## The output

**Key principle:** `dplyr` functions do **not** modify the original data; they return a *new* data frame

```{r}
starwars |>
  filter(species %in% c("Droid", "Human", "Ewok")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm=T))
```

**Question:** How do I save the output data so I can use it again?

## Assignment

```{r}
species_means <- starwars |>
  filter(species %in% c("Droid", "Human", "Ewok")) |>
  group_by(species) |>
  summarize(mean_height = mean(height, na.rm=T))

species_means

starwars
```

## Class activity

* Work with a neighbor on the class activity (handout)
* I will collect the handout at the end of class

## Activity

```{r, echo=F, message=F, warning=F}
library(Lahman)
library(tidyverse)

teams <- Teams |>
  select(yearID, franchID, W, L, attendance) |>
  filter(franchID %in% c("ANA", "BAL", "NYM"),
         yearID > 2020)
```

What is the `select` function doing?

```{r}
teams |>
  select(yearID, franchID, W, L)
```

## Activity

After running the following code, what will be the dimensions (number of rows and columns) of the `team_sub` dataset? Of the `teams` dataset?

```{r, eval=F}
team_sub <- teams |>
  filter(franchID %in% c("BAL", "NYM"))
```

## Activity

After running the following code, what will be the dimensions (number of rows and columns) of the `team_sub` dataset? Of the `teams` dataset?

```{r}
team_sub <- teams |>
  filter(franchID %in% c("BAL", "NYM"))

team_sub
```

## Activity

After running the following code, what will be the dimensions (number of rows and columns) of the `team_stats` dataset?

```{r, eval=F}
team_stats <- teams |>
  group_by(franchID) |>
  summarize(max_w = max(W), min_w = min(W))
```

## Activity

After running the following code, what will be the dimensions (number of rows and columns) of the `team_stats` dataset?

```{r}
team_stats <- teams |>
  group_by(franchID) |>
  summarize(max_w = max(W), min_w = min(W))

team_stats
```

## Activity

Give the output of the following code:

```{r, eval=F}
teams |>
  filter(franchID %in% c("BAL", "NYM")) |>
  group_by(franchID) |>
  summarize(max_attend = max(attendance))
```

## Activity

Give the output of the following code:

```{r}
teams |>
  filter(franchID %in% c("BAL", "NYM")) |>
  group_by(franchID) |>
  summarize(max_attend = max(attendance))
```

## Class activity

* Work with a neighbor on the class activity
* At the end of class, submit your work as an HTML file on Canvas (one per group, list all your names)

**Monday's class** will be reserved for getting Git and GitHub setup. We will use these tools for the rest of the semester.

* Before class, work through the **Git and GitHub assignment instructions** on the course website

