---
title: "Warmup: data wrangling"
format: pdf
editor: source
---

**Group members:**

**Instructions:** Work with a neighbor on the following activity. I will collect the handout at the end of class, and it will be part of your class participation grade. You will be graded only on effort -- it is ok if you don't finish all the questions, or get them all correct.

# Baseball data

The `teams` dataset contains information on wins, losses, and attendance for three MLB teams (Anaheim, Baltimore, and the New York Mets) between 2021 and 2023:

```{r, echo=F, message=F, warning=F}
library(Lahman)
library(tidyverse)

teams <- Teams |>
  select(yearID, franchID, W, L, attendance) |>
  filter(franchID %in% c("ANA", "BAL", "NYM"),
         yearID > 2020)
```

```{r}
teams
```


## Questions

1. The following chunk of code is run, producing the output below. What is the `select` function doing?

```{r, eval=F}
teams |>
  select(yearID, franchID, W, L)
```

```{r, echo=F}
teams |>
  select(yearID, franchID, W, L)
```

2. After running the following code, what will be the dimensions (number of rows and columns) of the `team_sub` dataset? Of the `teams` dataset?

```{r, eval=F}
team_sub <- teams |>
  filter(franchID %in% c("BAL", "NYM"))
```

\vspace{1cm}

3. After running the following code, what will be the dimensions (number of rows and columns) of the `team_stats` dataset?

```{r, eval=F}
team_stats <- teams |>
  group_by(franchID) |>
  summarize(max_w = max(W), min_w = min(W))
```

\vspace{1cm}

4. Give the output of the following code:

```{r, eval=F}
teams |>
  filter(franchID %in% c("BAL", "NYM")) |>
  group_by(franchID) |>
  summarize(max_attend = max(attendance))
```






