---
title: "Warmup: adult literacy data"
format: html
editor: source
---

**Instructions:** 

* Work with a neighbor to answer the following questions 
* You do *not* need to submit anything for this warmup activity

# Adult literacy data

The code below will load a dataset on historical adult literacy in different countries since 

```{r message=F, warning=F}
library(tidyverse)

litF <- read_csv("https://sta261-f26.github.io/class_activities/literacy_data.csv")
```

## Questions

1. What does a row in the data represent?

2. You want to fit a linear regression model of literacy rate on year, so you write the following code:
```{r, eval=F}
lm(literacy_rate ~ year, data = litF)
```
However, the code fails -- why?

3. You want to make the following scatterplot, of literacy rate vs. year. Can you create this plot with the `litF` data in its current form?

```{r, echo=F}
litF_long <- litF |>
  rename(country = starts_with("Adult")) |>
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "literacy_rate",
    values_drop_na = T
  )

litF_long |>
  ggplot(aes(x = year, y = literacy_rate)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Literacy rate") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

