---
title: "Activity: practice with R objects"
format: html
editor: source
---

## Practice with indexing

Consider the following vector `x`:

```{r}
x <- c("Statistics", "Mathematics", "Computer Science", "Spanish")
x
```

1. Use vector indexing to extract just "Statistics" and "Mathematics" from `x`.

**Solution:**

```{r}
x[1:2]
```


2. How could you extract the last entry in `x`, *without* manually counting the number of entries?

**Solution:**

```{r}
x[length(x)]
```


## More practice with indexing

Consider the following data frame `advisors`:

```{r}
advisors <- data.frame(name = c("Dr. Beavers",
                                "Dr. Huang",
                                "Dr. Evans",
                                "Dr. Mason",
                                "Dr. Alarcon"),
                       program = c("Statistics",
                                   "Statistics",
                                   "Statistics",
                                   "Math",
                                   "Spanish"),
                       office = c("MAN 337",
                                  "MAN 340",
                                  "MAN 329",
                                  "MAN 334",
                                  "GRN 519"))

advisors
```

3. Extract only the first and third columns of the `advisors` data frame, using indexing (`[ ]`) -- *not* the `select` function.

**Solution:**

```{r}
advisors[,c(1, 3)]
```

```{r}
advisors[c("name", "office")]
```



4. Now repeat question 3 with the `select` function instead.

**Solution:**

```{r, message=F, warning=F}
library(tidyverse)

advisors |>
  select(name, office)
```


5. Select only the rows of the `advisors` data frame corresponding to the statistics program, using indexing.

**Solution:**

```{r}
advisors[1:3,]
```

```{r}
advisors[advisors$program == "Statistics",]
```



6. Now repeat question 5 with the `filter` function instead.

**Solution:**

```{r}
advisors |>
  filter(program == "Statistics")
```



## Practice with R objects

The code below fits a linear regression model to the `diamonds` data:

```{r, message=F, warning=F}
m_diamonds <- lm(price ~ carat, data = diamonds)
```

7. What type of object is `m_diamonds` (i.e., how does R store the output of a fitted model?)

**Solution:** `m_diamonds` is a list

8. What information is contained in the `m_diamonds` object?

**Solution:** Coefficients, residuals, fitted values, and several other pieces of information about the model:

```{r}
names(m_diamonds)
```



## Practice with lists

9. Each of the following lists contains the vector `c(2, 7, 9)`. Determine the correct list indexing syntax to extract that vector from the list (your code should return just the vector, *not* a list containing the vector).

```{r}
x1 <- list(c(2, 7, 9))
x2 <- list(list(c(2, 7, 9)))
x3 <- list(c(2, 7, 9), list("a", "b"))
x4 <- list(c("a", "b"), list(list(c(2, 7, 9))))
```

**Solution:**

```{r}
x1[[1]]

x2[[1]][[1]]

x3[[1]]

x4[[2]][[1]][[1]]
```


10. Create a list `x5` such that the vector `c(2, 7, 9)` can be extracted with `x5[[3]][[2]][[1]]`.

**Solution:** Here is one possibility (there are many!)

```{r}
x5 <- list("a", "b", list("c", list(c(2, 7, 9))))
x5[[3]][[2]][[1]]
```

