---
title: "Activity: practice with R objects"
format: html
editor: source
---

**Instructions:** 

* Work with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta261-f26.github.io/class_activities/ca_09_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

## Practice with indexing

Consider the following vector `x`:

```{r}
x <- c("Statistics", "Mathematics", "Computer Science", "Spanish")
x
```

1. Use vector indexing to extract just "Statistics" and "Mathematics" from `x`.

2. How could you extract the last entry in `x`, *without* manually counting the number of entries?

## More practice with indexing

Consider the following data frame `advisors`:

```{r}
advisors <- data.frame(name = c("Dr. Beavers",
                                "Dr. Huang",
                                "Dr. Evans",
                                "Dr. Mason",
                                "Dr. Alarcon"),
                       program = c("Statistics",
                                   "Statistics",
                                   "Statistics",
                                   "Math",
                                   "Spanish"),
                       office = c("MAN 337",
                                  "MAN 340",
                                  "MAN 329",
                                  "MAN 334",
                                  "GRN 519"))

advisors
```

3. Extract only the first and third columns of the `advisors` data frame, using indexing (`[ ]`) -- *not* the `select` function.

4. Now repeat question 3 with the `select` function instead.

5. Select only the rows of the `advisors` data frame corresponding to the statistics program, using indexing.

6. Now repeat question 5 with the `filter` function instead.


## Practice with R objects

The code below fits a linear regression model to the `diamonds` data:

```{r, message=F, warning=F}
library(tidyverse)

m_diamonds <- lm(price ~ carat, data = diamonds)
```

7. What type of object is `m_diamonds` (i.e., how does R store the output of a fitted model?)

8. What information is contained in the `m_diamonds` object?


## Practice with lists

9. Each of the following lists contains the vector `c(2, 7, 9)`. Determine the correct list indexing syntax to extract that vector from the list (your code should return just the vector, *not* a list containing the vector).

```{r, eval=F}
x1 <- list(c(2, 7, 9))
x2 <- list(list(c(2, 7, 9)))
x3 <- list(c(2, 7, 9), list("a", "b"))
x4 <- list(c("a", "b"), list(list(c(2, 7, 9))))
```

10. Create a list `x5` such that the vector `c(2, 7, 9)` can be extracted with `x5[[3]][[2]][[1]]`.
