---
title: "Activity: Functions"
format: html
editor: source
---

**Instructions:** 

* Work with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta261-f26.github.io/class_activities/ca_17_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

## Computing sample variance

In statistics, we often summarize the variability or spread of a numeric variable by calculating the *sample variance*. Given $n$ observations $x_1,...,x_n$, the sample variance $s^2$ is defined by

$$s^2 = \frac{1}{n-1} \sum \limits_{i=1}^n (x_i - \overline{x})^2$$

In R, this can be done with the `var` function. For example:

```{r}
var(1:10)
```

1. **Question:** Here is an example vector:
```{r}
y <- c(1, 4, 8, 9, 21)
```
Using the definition above, compute the variance of the entries in `y`. Do not use the `var` function; rather, implement the mathematical definition I have provided. You may use other functions in R such as `sum` and `length`.


2. **Question:** Now write your own function to compute the sample variance, called `my_var`. You may use standard arithmetic operations in R, but do not use any existing implementations of the sample variance or standard deviation (e.g., don't use `var` or `sd` when writing your function).

```{r eval=F}
my_var <- function(x){
  # fill this in!
}
```


3. **Question:** Using your `my_var` function, and the `across` function, compute the variance for the all the numeric columns in the `diamonds` data.

## Trimmed means

One of the key lessons you learn in intro stats is that the mean is not a robust measure of center -- it is very susceptible to outliers! 
The median is one alternative that is more robust than the mean. Another alternative is a **trimmed mean**, in which we discard a proportion of data in either tail before averaging.

Suppose we have the following numbers (ordered from smallest to largest):

```
1  2  3  5  7  9 13 17 18 50
```

To compute the 10% trimmed mean, remove 10% of the observations in each tail (in this case, 1 observation from each tail), leaving us with:

```
2  3  5  7  9 13 17 18
```

The 10% trimmed mean is the mean of the remaining observations (in this case, the 10% trimmed mean is 8).

The 20% trimmed mean would remove 20% of the observations in each tail, etc. In general, the $p\cdot 100\%$ trimmed mean removes $p \cdot 100\%$ of the observations in each tail and then computes the mean of the remaining observations.

4. **Question:** Write a function called `trimmed_mean`, which takes two arguments: an input vector `x`, and a proportion `p`. The output of the function should be the $p \cdot 100\%$ trimmed mean of the entries in `x`.

```{r eval=F}
trimmed_mean <- function(x, p){
  # fill this in!
}
```

*Hints:* The `mean` and `quantile` functions may be helpful here!

5. **Question:** Compute the mean, median, and trimmed mean of the `carat` column in the diamonds data. How does the trimmed mean compare to the other two measures of center as you change $p$?

## 0-1 normalization

In class, we wrote a function called `z_score` to compute the z-score for each entry in a vector, by subtracting the mean of that vector and dividing by the standard deviation.

This is one example of *normalization*. Another type of normalization is 0-1 normalization. Given values $x_1,...,x_n$, the 0-1 normalized values are defined as

$$x_i^* = \frac{x_i - \min\{x_1,...,x_n\}}{\max\{x_1,...,x_n\} - \min\{x_1,...,x_n\}}$$

6. **Question:** Write a function called `zero_one_norm` to compute the 0-1 normalized value for each entry in a vector. Requirements:
    * Input: a numeric vector `x`
    * Output: a vector of the same length as `x`, containing the 0-1 normalized values
    
7. **Question:** Use your function to normalize each of the numeric columns in the `diamonds` data.


## Reflecting

8. Look back at each of the functions you have written. What assumptions are you making about the inputs to those functions? What would happen if the assumptions are not satisfied?



