---
title: "Exam 1 review"
format: html
editor: source
---

# Practice with data wrangling

## Reading data wrangling code in R

Here are two small datasets, `df1` and `df2`:

```{r, include=F}
library(tidyverse)

set.seed(3)

df1 <- data.frame(id = 1:5, 
                  x = sample(1:10, 5, replace=T),
                  y = sample(1:10, 5, replace=T),
                  z = sample(1:10, 5, replace=T))

df2 <- data.frame(id = 3:6,
                  a = sample(1:10, 4, replace=T),
                  b = sample(1:10, 4, replace=T))
```

```{r}
df1

df2
```


For each of the following chunks of code, write down the output or explain why it will cause an error.

1.

```{r}
df1 |>
  count(z)
```

2.

```{r}
df1 |>
  left_join(df2, join_by(id))
```

3.

```{r}
df1 |>
  inner_join(df2, join_by(id))
```

4.

```{r, error=TRUE}
df1 |>
  group_by(z) |>
  summarize(max_b = max(b))
```

Issue: `df1` has no column called `b`


5.

```{r, error = T}
df1 |>
  select(x, y) |>
  pivot_longer(cols = -id,
               names_to = "measurement",
               values_to = "value")
```

Issue: `id` doesn't appear in the data once we select on `x` and `y`

6.

```{r}
df1 |>
  select(id, x, y) |>
  pivot_longer(cols = -id,
               names_to = "measurement",
               values_to = "value") |>
  filter(id %in% c(1, 2, 3))
```

7.

```{r}
df1 |>
  left_join(df2, join_by(id)) |>
  mutate(new_var = x + a) |>
  group_by(z) |>
  summarize(mean_new_var = mean(new_var))
```


8.

```{r, error = T}
df1 |>
  left_join(df2, join_by(id)) |>
  mutate(new_var = x + a) |>
  group_by(z) |>
  summarize(mean_new_var = mean(new_var, na.rm=T)) |>
  summarize(mean_b = mean(b))
```

Issue: once we group and summarize, the only available variables are those used in the grouping and summaries

## Reading code in Python

Here is an example data frame, containing information on six members of a gym and their visits over the last month:

```{python}
#| echo: false
import pandas as pd
gym_members = pd.DataFrame({
  'id' : [1003, 248, 392, 40193, 5, 6923],
  'type' : ['student', 'adult', 'student', 'student', 'adult', 'senior'],
  'age': [18, 50, 20, 19, 50, 92],
  'visits': [10, 35, 2, 5, 5, 90],
  'fee': [50, 70, 50, 50, 70, 45]
  },
  index = ['a', 'b', 'c', 'd', 'e', 'f'])
```

```{python}
gym_members
```


For each of the following chunks of code, write down the output:

9. 

```{python}
gym_members.loc['c']
```

10.

```{python}
gym_members.iloc[3]
```

11.

```{python}
gym_members.groupby('type').agg(mean_age = ('age', 'mean'))
```

12.

```{python}
gym_members.assign(cost_per_visit = 
  gym_members['fee']/gym_members['visits'])
```

13.

```{python}
(gym_members.groupby('type')
           .agg(median_visits = ('visits', 'median'))
           .agg(max = ('median_visits', 'max')))
```

### Another example

Here are two data frames:

```{python}
#| include: false
patients = pd.DataFrame({
  'id': [24, 56, 11],
  'doctor': ['a', 'b', 'd'],
  'age': [33, 19, 20],
  'height': [155, 170, 160],
  'bp': [120, 115, 130]
})

providers = pd.DataFrame({
  'provider': ['a', 'b', 'c'],
  'location': ['Greensboro', 'Winston-Salem', 'Durham']
})
```

```{python}
patients
```

```{python}
providers
```



For each of the following chunks of code, write down the output. (Note: I am not going to take off points if the rows are in the wrong order).

14.

```{python}
patients.melt(id_vars = ['id', 'age'],
  value_vars = ['height', 'bp'],
  var_name = 'measurement',
  value_name = 'value')
```

15.

```{python}
patients.merge(providers, left_on = 'doctor', 
               right_on = 'provider', how = 'left')
```

16. 

```{python}
patients.merge(providers, left_on = 'doctor', 
               right_on = 'provider', how = 'inner')
```

### Another example

Here is another example:

```{python}
#| include: false
patients = pd.DataFrame({
  'id': [24, 56, 11],
  'doctor': ['a', 'b', 'd'],
  'age': [33, 19, 20],
  'bp_systolic': [120, 115, 130], 
  'bp_diastolic': [82, 83, 90]
})
```

```{python}
patients
```

Write the output of the following code:

17.

```{python}
pd.wide_to_long(patients, stubnames = 'bp',
  i = ['id', 'doctor', 'age'],
  j = 'type', sep = '_', suffix ='.+')
```



## Writing code

In each of the questions below, write code to produce the output from the original data.

18. Input data:


```{r, include=F}
library(tidyverse)
advisors <- data.frame(
  professor = c("Dr. Beavers", "Dr. Evans", "Dr. Hepler",
           "Dr. Mason", "Dr. Mason", "Dr. Esarey"),
  program = c("Statistics", "Statistics", "Statistics",
              "Math", "Psychology", "Politics"),
  office = c("MAN 337", "MAN 329", "MAN 341",
             "MAN 334", "GRN 438", "KRB 319")
)
```

```{r}
advisors
```


Output data in R:

```{r}
advisors |>
  count(program)
```

Output data in Python:

```{python}
#| include: false

advisors = r.advisors
```


```{python}
advisors.groupby('program').agg(n = ('office', 'count'))
```



19. Input data:

```{r, echo=F}
students <- data.frame(
  name = c("Ridwaan", "Coren", "Hallie", "Kathy", "Lianne"),
  major = c("Statistics", "Statistics", "Math", "Politics", "Statistics"),
  year = c(3, 2, 2, 4, 3),
  advisor = c("Dr. Beavers", "Dr. Evans", "Dr. Mason", "Dr. Esarey", "Dr. Hepler")
)
```

```{r}
students
```


Output data in R:

```{r}
students |>
  group_by(major) |>
  summarize(mean_year = mean(year))
```

Output data in Python:

```{python}
#| echo: false
students = r.students
```

```{python}
students.groupby('major').agg(mean_year = ('year', 'mean'))
```
20.

Input data:

```{r}
advisors
```

```{r}
students
```

Output data in R:

```{r}
students |>
  left_join(advisors, join_by(advisor == professor,
                              major == program))
```



Output data in Python:

```{python}
students.merge(advisors, left_on = ['advisor', 'major'],
               right_on = ['professor', 'program'],
               how = 'left').drop(['professor', 'program'],
                                             axis = 1)
```

**Note:** For the exam, it is ok if you don't do the `drop` piece

21. Input data:

```{r, include=F}
ex_df <- data.frame(
  group = c('a', 'b', 'c'),
  x_start = c(2, 3, 4),
  x_end = c(5, 6, 7),
  y_start = c(11, 12, 13),
  y_end = c(21, 22, 23)
)
```

```{r}
ex_df
```

Output data in R:

```{r}
ex_df |>
  pivot_longer(cols = -group,
               names_to = c(".value", "time"),
               names_sep = "_")
```

Output data in Python:

```{python}
#| echo: false
ex_df = r.ex_df
```

```{python}
pd.wide_to_long(ex_df, stubnames = ['x', 'y'], 
  i = 'group', j = 'time', sep = '_', suffix = ".+")
```

## Indexing in R 

22. Here is a vector in R. Write code to extract the third and fifth elements.


```{r, include=F}
x <- c(23, 48, 42, 19, 24, 67)
```

```{r}
x
```

```{r}
x[c(3, 5)]
```


23. Here is a list in R. Write code to extract the vector `2 6 1`:

```{r}
x <- list(1, 2, list(6, c(2, 6, 1), 1))
x
```

```{r}
x[[3]][[2]]
```


## Short answer

24. Describe some differences between data frames in R vs. pandas.

**Solution:** In terms of the objects themselves, the biggest difference is that DataFrames in pandas (like Series) always have an index. This means that there are multiple ways to access entries in a pandas DataFrame, and different data wrangling operations will interact differently with that index.

In terms of how we can use data frames when writing code, there are some other differences which are more reflective of differences between R and Python, as opposed to object specifics. For example, R supports tidy evaluation, which means that `df |> group_by(col1)` works because R knows that `col1` must refer to a column in the data frame. In pandas, those column names would always have to be in quotes. E.g. `df.groupby('col1')`

# Data analysis: water quality at Sydney beaches

High rainfall during summer 2025 caused concern about the water quality of beaches in Sydney, Australia.

To explore the relationship between rainfall and water quality, we have measurements on both water quality and weather data from 1991 to 2025.

**Research question:** How is rainfall related to E-coli bacteria levels?

## Data Dictionary

### `water_quality.csv`

|variable              |class     |description                           |
|:---------------------|:---------|:-------------------------------------|
|region                |character |Area of Sydney City |
|council               |character |City council responsible for water quality |
|swim_site             |character |Name of beach/swimming location |
|date                  |date      |Date |
|time                  |time      |Time of day |
|enterococci_cfu_100ml |integer   |Enterococci bacteria levels in colony forming units (CFU) per 100 millilitres of water |
|water_temperature_c   |integer   |Water temperature in degrees Celsius |
|conductivity_ms_cm    |integer   |Conductivity in microsiemens per centimetre |
|latitude              |double    |Latitude |
|longitude             |double    |Longitude |

### `weather.csv`

|variable         |class  |description                           |
|:----------------|:------|:-------------------------------------|
|date             |date   |Date |
|max_temp_C       |double |Maximum temperature in degrees Celsius |
|min_temp_C       |double |Minimum temperature in degrees Celsius |
|precipitation_mm |double |Rainfall in millimetres |
|latitude         |double |Latitude |
|longitude        |double |Longitude |

## Your task

Describe how you will use the available data to answer the research question. Your answer must include:

* A description of any data wrangling steps you will perform. For example, if you plan to create any new columns or compute summary statistics, describe how these new variables/statistics will be calculated. If you plan to combine data from multiple tables, on which variable(s) will you join? 
* An description of what the data will look like once you are done with any data wrangling (describe the rows and columns in your dataset).
* A description of at least one plot you will make -- sketch an example of what that plot could look like, with clear labels on the axes
* A description of at least one summary statistic or statistical model you will fit
* You do **not** need to write any code (although you may if you wish), but your answer should be sufficiently detailed that I could turn it into code


## Possible solution

When planning a data analysis, I think it helps to work backward from where we want to be. We want to know whether rainfall is related to E. coli bacteria levels.

This means that we need rainfall and E. coli levels as two different columns in the same dataset, with each row a location at a date and time. Our data should look something like this:

| swim site | date | time | enterococci_cfu_100ml | precipitation_mm | latitude | longitude | water_temperature_c | ... |
| --- | --- | --- | --- | --- | --- | --- | --- | --- |
| ... | ... | ... | ... | ... | ... | ... | ... | ... |



```{r, include=F}
water_quality <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-20/water_quality.csv')
weather <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-20/weather.csv')

comb_data <- water_quality |>
  inner_join(weather |> select(date, precipitation_mm),
             join_by(date))
```


We could then make a scatterplot, with E. coli levels on the y axis and precipitation on the x axis, and use correlation to describe the relationship.

To further investigate, I would suggest fitting a model, with E. coli as the response and precipitation as one of the explanatory variables. Since other factors may also be related to water quality, we would want to account for other explanatory variables (water temperature, water authority, etc.) in our model.

So how do we get there? We should join (I think an inner join would make the most sense, since we need both E. coli and precipitation information) `water_quality` and `weather` on date, and possibly on longitude and latitude, depending on how the data is recorded. If we want to join on longitude and latitude, we may need to identify the *closest* longitude and latitude, since the values may not exactly match between the two tables. 


